The Finnish company Next has a logistics problem in which used oils serve as raw material to form a class of renewable diesel. The supply chain team need to organise, to a minimal cost, the acquisition of two types of used oils (products p1 and p2) from three suppliers (supply nodes s1, s2, and s3) to feed the line of three of their used oils processing factories (demand nodes d1, d2, and d3). As the used oils are the byproduct of the supplier's core activities, the only requirement is that Next need to fetch the amount of oil and pay the transportation costs alone.

The oils have specific conditioning and handling requirements, so transportation costs vary between p1 and p2. Additionally, not all the routes (arcs) between suppliers and factories are available as some distances are not economically feasible. Table \ref{p1c1:tab:ex1-4_suplly_demand} shows the volume requirement of the two types of oil from each supply and demand node, and table \ref{p1c1:tab:ex1-4_arcs} show the transportation costs for each oil type per L and the arc capacity. Arcs with ``-'' for costs are not available as transportation routes.

\begin{table}[h!]
	\begin{subtable}[h]{0.4\textwidth}
		\begin{center}
		\begin{tabular}{c|cc}
			\textbf{node} & \textbf{p1} & \textbf{p2} \\
			\hline
			\textbf{s1 / d1} & 80 / 60 & 400 / 300 \\
			\textbf{s2 / d2} & 200 / 100 & 1500 / 1000 \\
			\textbf{s3 / d3} & 200 / 200 & 300 / 500 \\
		\end{tabular}
		\end{center}
		\caption{Supply availability and demand per oil type [in L]}
		\label{p1c1:tab:ex1-4_suplly_demand}
	\end{subtable}
	\hfill
	\begin{subtable}[h]{0.55\textwidth}
		\begin{center}
			\begin{tabular}{c|ccc}
				 & \textbf{d1} & \textbf{d2} & \textbf{d3}\\
				 & \textbf{p1/p2 (cap)} & \textbf{p1/p2 (cap)} & \textbf{p1/p2 (cap)}\\
				\hline
				\textbf{s1} & 5/- ($\infty$) & 5/18 (300) & -/- (0)\\
				\textbf{s2} & 8/15 (300) & 9/12 (700) & 7/14 (600)\\
				\textbf{s3} & -/- (0) & 10/20 ($\infty$) & 8/- ($\infty$)\\
			\end{tabular}
		\end{center}
		\caption{Arcs costs per oil type [in \euro \ per L] and arc capacities [in L]}
		\label{p1c1:tab:ex1-4_arcs}
	\end{subtable}
	\caption{Supply chain data}
\end{table}

Find the optimal oil acquisition plan for Next, i.e., solve its transportation problem to the minimum cost.
